---
title: "PS270 Final Project"
author: "Charlie Struhs"
description: "My final project"
output:
  distill::distill_article:
    self_contained: false
---




## Project thoughts

I am interested in exploring data related to conventions and whether or not the state turns to them when conventions are held there. I am also interested in looking at third party prevalence in America, such as where most people vote third party, what elections (national, gubernatorial, municipal) receive a bigger proportion of third party votes, and so on. Perhaps something even related to finance and politicians, where politicians choose to invest. Maybe something to do with federal spending and interest rates over time based on presidency and party. I am mainly concerned about how I will be able to formulate this as my skills in R are rudimentary, but I have no shortage of ideas as to what topics I would want to research and compile into a database.




https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&page=dataset&min-year=2&max-year=0&filter=0&year=2021

Does the size of government impact a nation's economic freedom? In this experiment, I plan to study how the sizes of governments influence a nation's ability to trade freely. My hypothesis is that a larger government, characterized by higher spending and greater regulatory control, will lead to more regulation on trade, and thus lower levels of economic freedom. This is grounded in the theory that larger governments often impose higher taxes and introduce more regulations which distort market functions, limit individual decision making, and reduce a nation's trade openness. Using data from the Fraser Institute on world economics, my explanatory variable will be variables such as size of government, which includes data regarding government consumption, subsidies, and state asset ownership. These indicators are scored on a scale from 0 to 10, 0 representing minimal government involvement and 10 maximum involvement. Other variables may include legal system efficiency, protection of property rights, and freedom to trade internationally, all of which may interact with government size to impact economic freedom. The dependent variable will be the overall economic freedom score, once again scaled from 0 to 10, 10 indicating the highest level of freedom. I expect to observe a negative correlation between government size and economic freedom. More importantly, a positive correlation between government size and economic freedom would disprove my hypothesis. 


Loading the data from Fraser Institute

```{r}
library(tidyverse)
library(readxl)
Fraser_data <- read_excel("Fraser_data****.xlsx")
Fraser_data
```
Key:

ISO Code <- 3 letter country code

countries <- countries

efi <- Economic Freedom Index, measured on a scale from 0 (least free) to 10 (most free)

rank <- rank out of all countries, 0 being highest efi, 165 being lowest

quartile <- 165 countries separated into quartiles, 0 being the highest quartile to 4 being the lowest quartile

govt_size <- size of government, 0 being smallest and 10 being largest

legal_rights <- legal rights, 0 being lowest to 10 being largest

sound_money <- sound currency indicator, 0 being not sound to 10 being most sound

intltrade_freedom <- Freedom to Trade internationally, 0 being low freedom to 10 being high freedom

regulation <- government regulation, 0 being regulations that deter entry into business, 10 being regulations that promote entry to business

If you would like more details regarding the Fraser Institute's approach, please click here :

https://www.fraserinstitute.org/economic-freedom/approach


 
```{r}
Fraser_data |>
ggplot(mapping = aes(x = govt_size, y = efi))+
       geom_point(color = "forestgreen")+
  labs(title = "Government Size vs. Economic Freedom Index",
       x = "Government Size", 
      y = "Economic Freedom Index",
      subtitle = "Source: Fraser Institute")

```

