---
title: "PS270 Final Project"
author: "Charlie Struhs"
description: "An Analysis of Government Size, Regulation, and Economic Freedom"
output:
  distill::distill_article:
    self_contained: false
---


Research Question and Main Hypothesis


Does the size of government impact a nation's economic freedom? In this experiment, I plan to study how the sizes of governments influence a nation's ability to trade freely and exercise economic freedom. My hypothesis is that larger governments-characterized by higher spending and greater regulatory control-will lead to more regulation on trade, leading to lower levels of economic freedom. This is rooted in the theory that larger governments often impose higher taxes and implement extensive regulations to increase government revenue, which distorts market functions, limits individual decision making, and reduces a nation's trade openness. 


Using data from the Fraser Institute on world economics (linked below), my explanatory variable will be the variable "size of government", which encompasses government consumption, subsidies, and state asset ownership. These indicators are scored on a scale from 0 to 10 for each country, 0 representing minimal government involvement and 10 maximum involvement. This means I will be measuring the entire population and do not require a random sample to estimate the population. The dependent variable will be the overall economic freedom score, once again scaled from 0 to 10, 10 indicating the highest level of freedom. I expect to observe a negative correlation between government size and economic freedom. That is, a larger government will have a lower economic freedom score. Conversely, a positive correlation between government size and economic freedom would disprove my hypothesis, meaning a larger government would have a higher economic freedom score. Another explanatory variable I will test will be regulation, which indicates ownership of banks, licensing restrictions, and other barriers to entry a government can impose on a market. This variable, also scored from 0-10, works differently than the government size score. A high score, in this case, indicates positive or well constructed regulations that open up market efficiency and increase competition. I predict a strong positive correlation between the regulation variable and the EFI. 


https://www.fraserinstitute.org/economic-freedom/dataset?geozone=world&page=dataset&min-year=2&max-year=0&filter=0&year=2021


Data

The data used in the analysis comes from the Fraser Institute on World Economics. Researchers at the Institute assign a score from 0-10 under 5 different categories: Size of government, legal system and security of property rights, sound money, freedom to trade internationally, and regulation. The dependent variable is the composite measure of economic freedom,the Economic Freedom Index, encompassing aspects of trade openness, property rights, and monetary policy. The independent variable is government size, which is measured through aspects such as government expenditures and taxes. Along with this, the regulation variable, which I will also test as an independent variable, assesses financial regulations and competition laws. The data was imported from an excel file and was edited in the file to show up nicely in Rstudio. One limitation, however, is the absence of some sub-saharan and asian countries that could provide a better glimpse into the economies of lesser developed nations. 

Data will be measured through a scatter plot model and a correlation will be derived. The data is cross sectional and analyzes countries at one point in time. Through two scatter plots comparing independent variables government size and regulation to the efi,  I hope to observe a strong negative correlation between government size and economic freedom, as this would prove my hypothesis, and a strong positive correlation between regulation and economic freedom. In this case, a strong negative correlation contains values of -0.75 or lower, while a strong positive correlation contains values of 0.75 or higher. between these values represents an insignificant correlation or weaker result.

 


Loading the data from Fraser Institute

```{r}
library(tidyverse)
library(readxl)
Fraser_data <- read_excel("Fraser_data****.xlsx")
Fraser_data
```
Key:

ISO Code <- 3 letter country code

countries <- countries

efi <- Economic Freedom Index, measured on a scale from 0 (least free) to 10 (most free)

rank <- rank out of all countries, 0 being highest efi, 165 being lowest

quartile <- 165 countries separated into quartiles, 0 being the highest quartile to 4 being the lowest quartile

govt_size <- size of government, 0 being smallest and 10 being largest

legal_rights <- legal rights, 0 being lowest to 10 being largest

sound_money <- sound currency indicator, 0 being not sound to 10 being most sound

intltrade_freedom <- Freedom to Trade internationally, 0 being low freedom to 10 being high freedom

regulation <- government regulation, 0 being regulations that deter entry into business, 10 being regulations that promote entry to business

If you would like more details regarding the Fraser Institute's approach, please click here :

https://www.fraserinstitute.org/economic-freedom/approach



Plotting data on a graph and calculating a correlation for government size and efi


```{r}
gs_efi_plot <- Fraser_data |>
ggplot(mapping = aes(x = govt_size, y = efi))+
       geom_point(color = "forestgreen", alpha = 0.6)+
  geom_smooth(color = "indianred") +
  labs(title = "Government Size vs. Economic Freedom Index",
       x = "Government Size", 
      y = "Economic Freedom Index",
      subtitle = "Source: Fraser Institute") +
  theme_minimal()
gs_efi_plot
  correlation1 <- cor(Fraser_data$efi, Fraser_data$govt_size, use = "pairwise")
  knitr::kable(correlation1, digits = 3, col.names = "correlation")
```
Analysis

It appears that as the government size increases in the initial stages, economic freedom is seen to increase at the fastest rate. Around the freedom rankings 6 and 7 there is a slight drop, then an increase, then finally a slow decrease as the government size approaches 10. Most results seem to be clustered between the rankings of 6 and 7. at the ranking of 6.5, the approximated efi by the geom_smooth() line is a little under 6.5.the nation with the highest efi, Hong Kong, has a government size of 7.99. The correlation value is 0.071, indicating a very weak, positive relationship between government size and economic freedom. Overall, the relationship between government size and EFI is hard to state. One could draw the conclusion that a lack or excessive amount of government size deters economic freedom in general. 

As an alternative approach, we can shift focus from government size to the regulation variable as the explanatory variable for the EFI. A higher score in government regulation indicates low barriers to entry, ability for corporations to determine wages, and the extent to which the banking industry is privately controlled. Examining government oversight on economic activity may give insight on excessive regulation and its ability to reduce competition and deter business entry, contributing to a lower EFI score.


Plotting data on a graph and calculating a correlation for government regulation and efi


```{r}
reg_efi_plot <- Fraser_data |>
ggplot(mapping = aes(x = regulation, y = efi))+
       geom_point(color = "forestgreen", alpha = 0.6)+
  geom_smooth(color = "indianred") +
  labs(title = "Government Regulation on Businesses vs. Economic Freedom Index",
       x = "Government Regulation", 
      y = "Economic Freedom Index",
      subtitle = "Source: Fraser Institute") +
  theme_minimal()
reg_efi_plot
 
correlation2 <- cor(Fraser_data$efi, Fraser_data$regulation, use = "pairwise")
knitr::kable(correlation2, digits = 3, col.names = "correlation")

```
The relationship here appears much clearer. The analysis reveals a strong positive correlation (a value of 0.897) between government regulation and the EFI, indicating that higher scores in government regulation are associated with greater economic freedom. This leads us to suggest that well designed regulatory measures may, in fact, enhance economic freedom by reducing barriers and increasing market transparency. This is contrary to the common assumption that regulation restricts economic freedom. Nations with higher regulation scores likely implement policies that enable free market activity rather than hinder it. 

Conclusion

Based on the findings, I reject my initial hypothesis that a larger sized government leads to a lack of economic freedom due to its insignificant correlation result. However, I maintain that regulatory controls do impact a nation's economic freedom seen by the strong positive correlation which proves my hypothesis in graph 2. My first explanatory variable showed that the size of a government does not necessarily indicate a nation's economic freedom. Rather, regulatory controls like price setting capabilities, presence of private banks and financial services, and laws supporting market competition are far more important. Limitations of the analysis include the cross-sectional nature of the data, meaning change over time is ignored in this study. Along with this, specific middle eastern and sub-saharan countries are ignored in the data, which could provide a stronger insight on economic freedom and perhaps regional variations in the data. Improvements to this study include collecting longitudinal data and comparing nation's scores over a period of time in the specific categories, in order to see economic freedom as ever changing rather than a static variable. 

Because I am measuring a population and do not require a random sample and hypothesis test generation, I am confident in the results of the correlation calculation as well as the graph data. A hypothesis test is unnecessary in this case because the parameters are known and we are not estimating from a sample. Nevertheless, our correlation values should not be interpreted causally. Further analysis, such as more controlled experiments, are necessary. Because the government size and regulation variables are hard to isolate and somewhat subjective, the correlation values show a strong relationship but should not be interpreted causally. 